
---
title: "HW06 Spatial Statistics"
author: "Doug Nychka"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Golden hourly ozone 2021

The goal of this activity is to build a functional linear model to forecast tomorrows hourly ozone from today's record. Keep in mind that although the basic model will be forecasting the basis coefficients our final goal is a good forecast of the 24 hour cycle of ozone for the next day.  Forecasts of the coefficients is really just an intermediate step. 

```{r}
suppressMessages(library( fields))
setwd("~/School/MATH498/Test2/")
```

First some setup based a complete data record where missing hours have been infilled using Tps. 

```{r}
tDay<- 1:365
load("O3CompleteExample.rda")
# SVD of data,  day by hour
lookSVD<- svd( O3Complete)
#print( lookSVD$d)
D<- diag(lookSVD$d)
# adjust signs to be more interpretable
# 
V<- -1*lookSVD$v
U<- -1*lookSVD$u
# weighted basis functions (columns)
VBasis<- V %*%D
```

Just for fun  and to get started here are the first four basis functions.

```{r}
fields.style()
matplot( 1:24, VBasis[,1:4], type="l", lty=1, lwd=2,
         xlab="hour", ylab="basis function")
title("Basis functions
      1 orange, 2 green, 3 blue, 4 red")
```

Here are the responses and lagged quantities to use both in terms of the hourly data and also in terms of the coefficients.
Note that we are trying forecast the full day's cycle: all 24 values of O3 for the next day. 


```{r}
N<- nrow(O3Complete )
tm<- tDay[-N] # can only forecast up to the second to last day!
Y<- O3Complete[-1,] # Tommorrows O3
X<-  O3Complete[-N,] # Todays O3
XU<- U[-N,] # todays coefficients
YU<- U[-1,] # tomorows coefficents
```

# Variance explained.
Based on the singular values we have the variance explained by each basis function and also the percent variance explained. We see the variance tends to decrease after 3 to 4 singular values. 

```{r}
print(  round(lookSVD$d^2,0) )
# and normalized as 
print(  100*cumsum(lookSVD$d^2)/ sum( lookSVD$d^2) )
```

This justifies using a fewer number of basis functions than 24!

# How *not* to proceed
If we just predicted every hour for the next day using the 24
ozone values from the previous day how many total parameters
would be in our forecast model?



# Getting started
How well do 4 basis functions approximate the full set of hourly values?  Recall that the approximation is 

```{r}
ApproxO3<- YU[,1:4]%*% t(VBasis[,1:4])
```

and compare to ```Y``` and report the RMSE for the difference. 
We will call this the *oracle* forecast because it assumes we can predict tomorrows coefficients perfectly. 


# forecast of first coefficient
Build an lm model to predict tommorows  first "U" coefficient, ```YU[,1]```,  based
on today's 4 "U" coefficients  ```XU[,1:4]```. 

Evaluate the forecast  RMSE by expanding with the first basis
function and compare to ```Y```. Compare this RMSE to how well
the *oracle* forecast RMSE.


# forecast of all 4 coefficients 

Now repeat using 4 lm models to forecast all 4 coefficients. 
Again how well does it do? 

# model checking 

Find the RMSE for your forecast for each day and plot these over time. Comment on any patterns. 

*Hint* use the apply function or **rowSums** to find the MSE for each day separately. 










